Log or not - Analysis of ECDC forecasts

### 2022-08-30 analysis/Analysis ECDC forecasts.Rmd
update paper draft

# situations with negative values: France 2021-05-22, CZ 2021-08-07, ES 2021-03-06
```

```{r}
# helper functions
# ------------------------------------------------------------------------------

@@

@@

```{r}
# load data 
# ------------------------------------------------------------------------------
hub_data <- rbindlist(list(
  fread(here("data", "full-data-european-forecast-hub-1.csv")), 
  fread(here("data", "full-data-european-forecast-hub-2.csv")),
  fread(here("data", "full-data-european-forecast-hub-3.csv"))
)) |>
  unique()
```

@@

```{r}
scores <- fread(here("output", "data", "all-scores-european-hub.csv"))
scores[, scale := factor(scale, levels = c("natural", "log"))]
data.table::setnames(scores, old = c("aem", "sharpness"), new = c("ae_median", "dispersion"))

scores_gr <- fread(here("output", "data", "all-scores-growth-rate-european-hub.csv"))
scores_gr[, scale := factor(scale, levels = c("natural", "log"))]
```

@@

plot_means_obs <- mean_observations |>
  ggplot(aes(y = mean_value, x = reorder_within(location, -mean_value, target_type))) +   geom_bar(stat = "identity", fill = "grey50") + 
  facet_wrap(~ target_type, scales = "free") + 
  theme_scoringutils() + 
  scale_y_continuous(labels = label_fn) + 
  scale_x_discrete(guide = guide_axis(n.dodge=2), labels = label_fn_within) + 
  labs(y = "Mean observed value", x = "Location")

box_plot_obs <- hub_data |>
  mutate(true_value = pmax(true_value, 1)) |>
  dplyr::select(location, target_type, target_end_date, true_value) |>
  unique() |>
  ggplot(aes(y = true_value, x = target_type)) + 
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") + 
  labs(y = "Observations", x = "Target type")

plot_df <- scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |> 
  full_join(mean_observations)

plot_mean_scores <- plot_df |>
  ggplot(aes(y = interval_score,
             x = reorder_within(location, -mean_value, target_type))) + 
  geom_bar(stat = "identity") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(scale ~ target_type, scale = "free") + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn) + 
  scale_x_discrete(guide = guide_axis(n.dodge=2), labels = label_fn_within) +
  labs(y = "Mean interval score", x = "Location")

@@

box_plot_horizon <- scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon <= 4) |>
  group_by(forecast_date, model, location, target_type, scale) |>
  mutate(interval_score = interval_score / interval_score[horizon == 1]) |>
  ggplot(aes(y = interval_score,
             x = as.factor(horizon), fill = target_type)) + 
  geom_hline(aes(yintercept = 1), linetype = "dashed", color = "grey40") +
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) +
  geom_boxplot(alpha = 0.5, position = "dodge2") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ scale, nrow = 1) + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") +
  labs(y = "Rel. increase in WIS", x = "Forecast horizon (weeks)")

layout <- "
AB
CD
CD
EE"

plot_means_obs  + box_plot_obs + plot_mean_scores + box_plot_scores + box_plot_horizon +
  plot_layout(heights = c(1, 2, 1), 
              widths = c(3, 1)) +
  plot_layout(design = layout) +
  plot_annotation(tag_levels = "A")

@@

@@

ggsave("output/figures/HUB-correlations.png", width = 7, height = 6)

@@

```{r}
summarised_pairwise <- scores |>
  filter(target_type == "Cases", 
         horizon == 2) |>
  summarise_scores(by = c("model", "scale"), relative_skill = TRUE)

ranking_natural <- 
  summarised_pairwise |>
  filter(scale == "natural") |>
  arrange(-relative_skill) |>
  pull(model)

plot_rel_skill <- summarised_pairwise |>
  mutate(model = factor(model, levels = ranking_natural)) |>
  ggplot(aes(y = model, x = relative_skill)) +
  geom_bar(stat = "identity") +
  theme_scoringutils() +
  facet_wrap(~scale) +
  labs(y = "Model", x = "Relative skill")

diffs <- summarised_pairwise |>
  mutate(model = factor(model, levels = ranking_natural)) |>
  select(model, relative_skill, scale) |>
  pivot_wider(names_from = scale, values_from = relative_skill) |>
  mutate(model = factor(model, levels = ranking_natural), 
         difference = log - natural)

plot_diffs <- diffs |>
  mutate(title = "difference") |>
  ggplot(aes(y = model, x = difference)) +
  geom_bar(stat = "identity") +
  theme_scoringutils() +
  labs(y = "Model", x = "Difference in relative skill") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank()) +
  facet_wrap(~title)

# plot_distr_scores <- scores |>
#   filter(horizon == 2) |>
#   mutate(model = factor(model, levels = ranking_natural)) |>
#   ggplot(aes(y = model, x = interval_score)) +
#   ggdist::stat_pointinterval() +
#   facet_wrap(~ scale, scale = "free_x") +
#   theme_scoringutils() +
#   scale_x_log10(labels = label_fn) +
#   theme(axis.title.y = element_blank(), 
#         axis.text.y = element_blank(), 
#         axis.ticks.y = element_blank(), 
#         axis.line.y = element_blank()) +
#   labs(y = "Model", x = "WIS")

plot_rel_wis <- scores |>
  filter(horizon == 2, 
         target_type == "Cases") |>
  mutate(model = factor(model, levels = ranking_natural)) |>
  summarise_scores(by = c("model", "scale")) |>
  plot_wis(relative_contributions = FALSE) +
    theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank()) +
  scale_x_continuous(labels = label_fn) +
  facet_wrap(~scale, scales = "free_x")

plot_ranking_change <- summarised_pairwise |>
  select(model, scale, relative_skill) |>
  group_by(scale) |>
  mutate(rank = rank(-relative_skill), 
         x = ifelse(scale == "natural", 0.4, 2.8), 
         x_arrow = ifelse(scale == "natural", 1.7, 2.7)) |>
  group_by(model) |>
  mutate(color = ifelse(diff(relative_skill) > 0, "deteriorated", "not deteriorated")) |>
  ggplot(aes(y = rank, x = x, group = model, label = model)) +
    geom_path(aes(x=x_arrow, color = color), arrow = arrow(length = unit(0.02,"npc")), 
            size=1, show.legend = FALSE) +
  geom_text(size=3.5, hjust = 0) +
  coord_cartesian(xlim = c(0.5, 4)) +
  theme_scoringutils() +
  theme(axis.title = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(), 
        legend.position="none")

layout <- "
AAA
BCD"

plot_ranking_change + plot_rel_skill + plot_diffs + plot_rel_wis + #plot_distr_scores 
  plot_layout(design = layout) +
  plot_layout(heights = c(1, 2)) +
  plot_layout(widths = c(2, 1, 2)) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave("output/figures/HUB-pairwise-comparisons.png", width = 9, height = 5)

@@

>>>>

# situations with negative values: France 2021-05-22, CZ 2021-08-07, ES 2021-03-06

# helper functions
# ------------------------------------------------------------------------------

@@

# load data 
# ------------------------------------------------------------------------------
hub_data <- rbindlist(list(
  fread(here("data", "full-data-european-forecast-hub-1.csv")), 
  fread(here("data", "full-data-european-forecast-hub-2.csv")),
  fread(here("data", "full-data-european-forecast-hub-3.csv"))
)) |>
  unique()

scores <- fread(here("output", "data", "all-scores-european-hub.csv"))
scores[, scale := factor(scale, levels = c("natural", "log"))]
data.table::setnames(scores, old = c("aem", "sharpness"), new = c("ae_median", "dispersion"))

scores_gr <- fread(here("output", "data", "all-scores-growth-rate-european-hub.csv"))
scores_gr[, scale := factor(scale, levels = c("natural", "log"))]

@@

@@

@@

plot_means_obs <- mean_observations |>
  ggplot(aes(y = mean_value, x = reorder_within(location, -mean_value, target_type))) +   geom_bar(stat = "identity", fill = "grey50") + 
  facet_wrap(~ target_type, scales = "free") + 
  theme_scoringutils() + 
  scale_y_continuous(labels = label_fn) + 
  scale_x_discrete(guide = guide_axis(n.dodge=2), labels = label_fn_within) + 
  labs(y = "Mean observed value", x = "") +
  theme(axis.title.x = element_blank())

box_plot_obs <- hub_data |>
  mutate(true_value = pmax(true_value, 1)) |>
  dplyr::select(location, target_type, target_end_date, true_value) |>
  unique() |>
  ggplot(aes(y = true_value, x = target_type)) + 
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") + 
  labs(y = "Observations", x = "")  +
  theme(axis.title.x = element_blank())

plot_df <- scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale, type_and_scale) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |> 
  full_join(mean_observations)

plot_mean_scores <- plot_df |>
  ggplot(aes(y = interval_score,
             x = reorder_within(location, -mean_value, target_type))) + 
  geom_bar(stat = "identity") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ type_and_scale, scale = "free") + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn) + 
  scale_x_discrete(guide = guide_axis(n.dodge=2), labels = label_fn_within) +
  labs(y = "Mean interval score", x = "Loaction")

@@

box_plot_horizon <- scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon <= 4) |>
  group_by(forecast_date, model, location, target_type, scale) |>
  mutate(interval_score = interval_score / interval_score[horizon == 1]) |>
  ggplot(aes(y = interval_score,
             x = as.factor(horizon), fill = target_type)) + 
  geom_hline(aes(yintercept = 1), linetype = "dashed", color = "grey40") +
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) +
  geom_boxplot(alpha = 0.5, position = "dodge2") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ scale, nrow = 1) + 
  theme_scoringutils() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") +
  labs(y = "Rel. change in WIS", x = "Forecast horizon (weeks)")

layout <- "
AB
CD
CD
EF"

plot_means_obs  + box_plot_obs + plot_mean_scores + box_plot_scores + box_plot_horizon + plot_spacer() +
  plot_layout(heights = c(1, 2, 1), 
              widths = c(3, 1)) +
  plot_layout(design = layout) +
  plot_annotation(tag_levels = "A")

@@

p_cor_skill <- correlation_rel_skill |>
  rename(`Target type` = target_type) |>
  ggplot(aes(x = horizon, y = cor, fill = `Target type`, color = `Target type` )) + 
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Cor(rel. skill)", x = "Forecast horizon")

@@

ggsave("output/figures/HUB-correlations.png", width = 7, height = 5.5)

@@

ranking_figure <- function(target = "Cases") {
  summarised_pairwise <- scores |>
  filter(target_type == target, 
         horizon == 2) |>
  summarise_scores(by = c("model", "scale"), relative_skill = TRUE)

ranking_log <- 
  summarised_pairwise |>
  filter(scale == "log") |>
  arrange(-relative_skill) |>
  pull(model)

plot_rel_skill_natural <- summarised_pairwise |>
  filter(scale == "natural") |>
  ggplot(aes(y = reorder(model, -relative_skill), x = relative_skill)) +
  geom_bar(stat = "identity") +
  theme_scoringutils() +
  facet_wrap(~scale) +
  labs(y = "Model", x = "Relative skill")

plot_rel_skill_log <- summarised_pairwise |>
  filter(scale == "log") |>
  ggplot(aes(y = reorder(model, -relative_skill), x = relative_skill)) +
  geom_bar(stat = "identity") +
  theme_scoringutils() +
  facet_wrap(~scale) +
  labs(y = "", x = "Relative skill") +
  theme(axis.title.y = element_blank())

diffs <- summarised_pairwise |>
  select(model, relative_skill, scale) |>
  pivot_wider(names_from = scale, values_from = relative_skill) |>
  mutate(model = factor(model, levels = ranking_log), 
         difference = log - natural)

plot_diffs <- diffs |>
  mutate(title = "difference") |>
  ggplot(aes(y = model, x = difference)) +
  geom_bar(stat = "identity") +
  theme_scoringutils() +
  labs(y = "Model", x = "\u0394 Relative skill") +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank()) +
  facet_wrap(~title)

plot_rel_wis <- scores |>
  filter(horizon == 2, 
         target_type == "Cases") |>
  mutate(model = factor(model, levels = ranking_log)) |>
  summarise_scores(by = c("model", "scale")) |>
  plot_wis(relative_contributions = TRUE) +
    theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank()) +
  scale_x_continuous(labels = label_fn) +
  facet_wrap(~scale, scales = "free_x")

plot_ranking_change <- summarised_pairwise |>
  select(model, scale, relative_skill) |>
  group_by(scale) |>
  mutate(rank = rank(-relative_skill), 
         x = ifelse(scale == "natural", 0.4, 2.8), 
         x_arrow = ifelse(scale == "natural", 1.7, 2.7)
         ) |>
  group_by(model) |>
  mutate(color = ifelse(diff(relative_skill) > 0, "deteriorated", "not deteriorated")) |>
  ggplot(aes(y = rank, x = x, group = model, label = model)) +
    geom_path(
      aes(x=x_arrow, color = color), 
      arrow = arrow(length = unit(0.09,"npc")), 
      lineend = "round", linejoin = "mitre",
      size=1, show.legend = FALSE) +
  # geom_text(size=3.3, hjust = 0) +
  coord_cartesian(#xlim = c(0.5, 4), 
                  ylim = c(0.8, 9.15)) +
  theme_scoringutils() +
  theme(axis.title = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(), 
        legend.position="none")

# layout <- "
# AAA
# BCD"

return(
  list(
    plot_ranking_change, plot_rel_skill_natural, plot_rel_skill_log, plot_diffs, plot_rel_wis
  )
)

plot_ranking_change + plot_rel_skill + plot_diffs + plot_rel_wis + #plot_distr_scores 
  plot_layout(design = layout) +
  plot_layout(heights = c(1, 2)) +
  plot_layout(widths = c(2, 1, 2)) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
}

ranking_cases <- ranking_figure(target = "Cases")
ranking_deaths <- ranking_figure(target = "Deaths")

layout <- "
ABCDE
FGHIJ"

ranking_cases[[2]] + ranking_cases[[1]] + 
ranking_cases[[3]] + ranking_cases[[4]] + ranking_cases[[5]] + 
  ranking_deaths[[2]] + ranking_deaths[[1]] + 
ranking_deaths[[3]] + ranking_deaths[[4]] + ranking_deaths[[5]] +
  plot_annotation(tag_levels = "A") +
  plot_layout(design = layout, 
              guides = "collect") &
  theme(legend.position = "bottom")

ggsave("output/figures/HUB-pairwise-comparisons.png", width = 10.5, height = 6)

@@

ensemble_cases <- plot_pred_score(hub_data |> filter(location == "DE"), 
                                  scores, "epiforecasts-EpiNow2", 
                                  remove_x_text = TRUE, 
                                  type = "Cases")

ensemble_deaths <- plot_pred_score(hub_data |> filter(location == "DE"), 
                                   scores, "EuroCOVIDhub-baseline", 
                                   type = "Deaths")
